---
title: "EDA Executive Summary - WHO World Health Statistics 2021"
subtitle: "Data Science I (STAT 301-1)"
author: "Chloe Sokol"
output:
  html_document:
    toc: true
    toc_float: true
    highlight: "tango"
    code_folding: show
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE, comment = FALSE, message = FALSE
)
```

## Overview
The data set is sourced from the World Health Organization and represents WHO’s annual World Health Statistics report for the 194 WHO member states in 2021, though most of the data falls under a previous year. The dataset was found on the WHO website within the Global Health Observatory. For 2021, they have a full report as well as a visual summary in addition to the raw .xls file. (https://www.who.int/data/gho/publications/world-health-statistics)

```{r, echo=FALSE}
# Packages
library(tidyverse)
library(ggplot2)
library(readxl)
library(viridis)
library(patchwork)

# Dataset
whs <- read_excel("data/processed/whs2021.xlsx")
```

## Executive Summary

The most interesting finding of the long-form EDA was an exploration of the data on vaccine coverage in the WHS dataset. Three immunizations covered by the dataset for young children include Diphtheria-tetanus-pertussis (DTP3) and Pneumococcal conjugate 3rd dose (PCV3) among 1-year-olds, as well as the Measles-containing-vaccine second-dose (MCV2) among the nationally recommended age which is still during young childhood. While the data on each country, region, and global statistics for each vaccine was compiled in the 2021 World Health Statistics Report, observations for these three variables were recorded in 2019.
<br><br>
The region with the highest DTP3 immunization coverage among 1-year-olds is the European region (95%), followed by the Western Pacific region (94%), the Southeast Asia region (91%), the region of the Americas (84%), the Eastern Mediterranean region (82%), and finally, the African region (74%). For the MCV2 immunization coverage by the nationally recommended age, the European region (92%) had the highest proportion, followed by the Western Pacific region (91%), the Southeast Asia region (83%), the Eastern Mediterranean region and the region of the Americas (75%), and finally, the African region (33%). Finally, the region with the highest PCV3 immunization coverage among 1-year olds is the region of the Americas (83%), followed by the European region (80%), the African region (70%), the Eastern Mediterranean region (52%), the Southeast Asia region (23%), and finally, the Western Pacific region (14%). This data is shown in the faceted bar chart below.
<br><br>
```{r, echo=FALSE}
regions <- whs %>%
  filter(str_detect(member_state, "Region")) %>%
  mutate(DTP3 = DTP319,
         MCV2 = MCV219,
         PCV3 = PCV319)

immune_stat <- gather(regions, key="measure", value="value", c("DTP3", "MCV2", "PCV3"))

ggplot(immune_stat, aes(x=member_state, y=value))+
  geom_bar(stat='identity', aes(fill = factor(member_state))) +
  facet_wrap(~ measure) +
  coord_flip() +
  theme(legend.position = "none") +
  scale_fill_viridis_d(option = "inferno") +
  labs(title = "Immunization Coverage") +
  ylab("Percent") + xlab("Region")
```
```{r, echo=FALSE}
plot1 <- whs %>%
  ggplot(aes(x = DTP319, y = u5_mortality19)) +
  geom_point() + 
  labs(title = "DTP3 Immunization") +
  xlab("% Coverage (1yr)") + ylab("Under 5 Mortality Ratio (Per 1000 Live Births)")

plot2 <- whs %>%
  ggplot(aes(x = MCV219, y = u5_mortality19)) +
  geom_point() + 
  labs(title = "MCV2 Immunization") +
  xlab("% Coverage (Rec. Age)") + ylab("Under 5 Mortality Ratio (Per 1000 Live Births)")

plot3 <- whs %>%
  ggplot(aes(x = PCV319, y = u5_mortality19)) +
  geom_point() + 
  labs(title = "PCV3 Immunization") +
  xlab("% Coverage (1yr)") + ylab("Under 5 Mortality Ratio (Per 1000 Live Births)")

plot1 + plot2 + plot3
```
<br><br>
The order, in terms of regions with the highest vaccination coverage for diphtheria-tetanus-pertussis and measles-containing-vaccine second-dose in their respective age groups was almost exactly the opposite order of regions for under 5 mortality rates (per 1000 live births), in which rates highest to lowest were seen in the African region (74), followed by the Eastern Mediterranean region (46), the Southeast Asia region (32), the region of the Americas (13), the Western Pacific region (11), and finally, the European region (8). This reciprocal observation is not seen for pneumococcal conjugate 3rd dose immunization coverage among 1-year olds. Furthermore, there is a weekly negative relationship between vaccination coverage for all three vaccines and the neonatal mortality rate. While the relationship can be rationalized as young children vaccinated against life-threatening diseases are less likely to die in the under 5 age range, we cannot make any causal claims from the data due to the potential for confounders.